---
title: "Data preprocessing"
date: "2025-03-04"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

# Overview

Generate some figures and clean up publicly available data for course material 

1. Clean up data
2. Generate figures


# Pre-processing

```{r}
library(ggplot2)
library(gridExtra)
library(biomaRt)
library(dplyr)
```


# Objective 1: Clean up public data 

We obtain the data from here - https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-GEOD-50760. Also using these tutorials ;

- https://genviz.org/module-04-expression/0004/02/01/DifferentialExpression/
- https://bioinformatics.ccr.cancer.gov/btep/courses/bioinformatics-for-beginners-b4b-introduction-to-rna-sequencing
- https://hbctraining.github.io/Intro-to-R/
- https://github.com/hbctraining/DGE_workshop?tab=readme-ov-file


We want to clean the data to make it so that ;

- 20 patients
  - 15 normal
  - 15 cancer
- About 20,000 genes

## Raw counts

Lets read in actual data from an experiment comparing tumor and normal

Read in counts

```{r}
rawCounts <- read.delim("http://genomedata.org/gen-viz-workshop/intro_to_deseq2/tutorial/E-GEOD-50760-raw-counts.tsv")
```

Annotate which are protein coding

```{r}
# Use Ensembl as the BioMart database
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

# Example gene list (change this to your list)
genes <- rawCounts$Gene.Name # mix of coding and non-coding

# Query for gene type
annot <- getBM(
  attributes = c("hgnc_symbol", "ensembl_gene_id", "gene_biotype"),
  filters = "hgnc_symbol",
  values = genes,
  mart = ensembl
)
```

```{r}
df_gene <- rawCounts %>% 
  dplyr::select(Gene.ID, Gene.Name) %>% 
  dplyr::rename(ensembl_gene_id = Gene.ID) %>% 
  left_join(annot) %>% 
  dplyr::filter(gene_biotype == "protein_coding") %>% 
  dplyr::filter(!is.na(Gene.Name)) %>% 
  distinct(Gene.Name, .keep_all = TRUE)
```

Subset the raw counts to these genes

```{r}
rawCounts_sub <- rawCounts %>% 
  dplyr::filter(Gene.ID %in% df_gene$ensembl_gene_id)
```

## Metadata

Read in sample mapping

```{r}
# Read in the sample mappings
sampleData <- read.delim("http://genomedata.org/gen-viz-workshop/intro_to_deseq2/tutorial/E-GEOD-50760-experiment-design.tsv")
```



# Objective 2: Figure generation

## Random cells of tumor normal

We just want to do a simple comparison, comparing tumor to normals.

```{r}
# Create a visualization with balls representing normal vs cancer cells
# Create data for normal cells (more organized pattern)
set.seed(123)
normal_cells <- data.frame(
  x = runif(50, 0, 10),
  y = runif(50, 0, 10),
  group = "Normal"
)

# Create data for cancer cells (more scattered pattern)
cancer_cells <- data.frame(
  x = runif(50, 0, 10),
  y = runif(50, 0, 10),
  group = "Cancer"
)

# Combine the data
all_cells <- rbind(normal_cells, cancer_cells)

# Create the plot
ggplot(all_cells, aes(x = x, y = y, color = group)) +
  geom_point(size = 4, alpha = 0.7) +
  facet_wrap(~ group) +
  theme_minimal() +
  labs(title = "Cell Organization: Normal vs Cancer",
       x = "", y = "") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_color_manual(values = c("Normal" = "skyblue", "Cancer" = "salmon")) +
  coord_fixed()
```

## Five genes with read counts

```{r}
# Create data for 5 genes
genes <- paste0("Gene", 1:5)
n_reads <- 10  # base number of reads

# Create data frame for normal samples with differential expression
normal_counts <- c(10, 8, 12, 15, 7)  # Different counts for each gene
normal_data <- data.frame(
  gene = rep(genes, times = normal_counts),
  x = runif(sum(normal_counts), 0.2, 0.8),
  y = rep(5:1, times = normal_counts) + runif(sum(normal_counts), -0.3, 0.3),
  sample = "Normal"
)

# Create data frame for tumor samples with differential expression
tumor_counts <- c(5, 12, 8, 7, 15)  # Different counts for each gene
tumor_data <- data.frame(
  gene = rep(genes, times = tumor_counts),
  x = runif(sum(tumor_counts), 0.2, 0.8),
  y = rep(5:1, times = tumor_counts) + runif(sum(tumor_counts), -0.3, 0.3),
  sample = "Tumor"
)

# Combine the data
all_data <- rbind(normal_data, tumor_data)

# Create summary data for bar plot
summary_data <- data.frame(
  gene = rep(genes, 2),
  sample = rep(c("Normal", "Tumor"), each = 5),
  count = c(normal_counts, tumor_counts)
)

# Create the main plot
main_plot <- ggplot() +
  # Add gene rectangles
  geom_rect(data = data.frame(ymin = 0.5:4.5, ymax = 1.5:5.5, 
                             xmin = 0, xmax = 1),
            aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax),
            fill = "grey90", color = "black") +
  # Add read circles
  geom_point(data = all_data, 
             aes(x = x, y = y, color = sample),
             size = 3, alpha = 0.7) +
  # Facet by sample type
  facet_wrap(~ sample, nrow = 1) +
  # Add gene labels
  scale_y_continuous(breaks = 1:5, labels = rev(genes)) +
  # Customize theme
  theme_minimal() +
  labs(title = "Read Distribution Across 5 Genes",
       y = "Genes", x = "") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 12, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_color_manual(values = c("Normal" = "skyblue", "Tumor" = "salmon")) +
  xlim(0, 1)

main_plot
# # Create the bar plot
# bar_plot <- ggplot(summary_data, aes(x = gene, y = count, fill = sample)) +
#   geom_bar(stat = "identity", position = "dodge", width = 0.7) +
#   scale_fill_manual(values = c("Normal" = "skyblue", "Tumor" = "salmon")) +
#   theme_minimal() +
#   labs(x = "", y = "Read Count") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1),
#         panel.grid = element_blank(),
#         legend.position = "none") +
#   ylim(0, max(c(normal_counts, tumor_counts)) * 1.1)
# 
# # Combine the plots
# grid.arrange(main_plot, bar_plot, ncol = 2, widths = c(2, 1))
```

## Generate 20 samples

For 10 samples

```{r}
# Create data for 5 genes
genes <- paste0("Gene", 1:5)
n_samples <- 10  # number of samples per group

# Create sample IDs
sample_ids <- paste0(rep(c("Normal", "Tumor"), each = n_samples), "_", 1:n_samples)

# Create base expression levels for each gene in normal samples
normal_base <- c(10, 8, 12, 15, 7)

# Create base expression levels for each gene in tumor samples
tumor_base <- c(5, 12, 8, 7, 15)

# Create data frame with all samples
all_data <- data.frame()

# Generate data for each sample with some random variation
for(i in 1:n_samples) {
  # Normal samples
  normal_counts <- round(normal_base * runif(5, 0.8, 1.2))  # Add 20% random variation
  normal_data <- data.frame(
    gene = rep(genes, times = normal_counts),
    x = runif(sum(normal_counts), 0.2, 0.8),
    y = rep(5:1, times = normal_counts) + runif(sum(normal_counts), -0.3, 0.3),
    sample = paste0("Normal_", sprintf("%02d",i))
  )
  
  # Tumor samples
  tumor_counts <- round(tumor_base * runif(5, 0.8, 1.2))  # Add 20% random variation
  tumor_data <- data.frame(
    gene = rep(genes, times = tumor_counts),
    x = runif(sum(tumor_counts), 0.2, 0.8),
    y = rep(5:1, times = tumor_counts) + runif(sum(tumor_counts), -0.3, 0.3),
    sample = paste0("Tumor_", sprintf("%02d",i))
  )
  
  all_data <- rbind(all_data, normal_data, tumor_data)
}

# Create summary data for bar plot
summary_data <- data.frame(
  gene = rep(genes, 2),
  group = rep(c("Normal", "Tumor"), each = 5),
  count = c(normal_base, tumor_base)
)

# Create the main plot
main_plot <- ggplot() +
  # Add gene rectangles
  geom_rect(data = data.frame(ymin = 0.5:4.5, ymax = 1.5:5.5, 
                             xmin = 0, xmax = 1),
            aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax),
            fill = "grey90", color = "black") +
  # Add read circles
  geom_point(data = all_data, 
             aes(x = x, y = y, color = sample),
             size = 2, alpha = 0.5) +
  # Facet by sample
  facet_wrap(~ sample, nrow = 2) +
  # Add gene labels
  scale_y_continuous(breaks = 1:5, labels = rev(genes)) +
  # Customize theme
  theme_minimal() +
  labs(title = "Read Distribution Across 5 Genes in 20 Samples",
       y = "Genes", x = "") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 8),
        legend.position="none",
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_color_manual(values = c(
    rep("skyblue", n_samples),  # Normal samples
    rep("salmon", n_samples)    # Tumor samples
  )) +
  xlim(0, 1)

# Create the bar plot
bar_plot <- ggplot(summary_data, aes(x = gene, y = count, fill = group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Normal" = "skyblue", "Tumor" = "salmon")) +
  theme_minimal() +
  labs(x = "", y = "Average Read Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.position = "none") +
  ylim(0, max(c(normal_base, tumor_base)) * 1.1)

# Combine the plots
grid.arrange(main_plot, bar_plot, ncol = 2, widths = c(3, 1))
```
